---
title: "R Notebook_NoiseproR"
author: Aliff Firdaus Mat Yusoff
Student ID: KGJ 140066
output: html_notebook
---

# Introduction

This R notebook serves to record statistical analysis process (i.e., method, consideration, R script) for noise exposure profiling using R. The objectives are to facilitate review of all statistical method adopted as well as learning to ensure reproducibility of this study in the future.


# Pre-requisite

The following packages need to be installed in order to run the codes provided in this document:
```{r}
install.packages("dplyr")
install.packages("tidyr")
install.packages("ggplot2")
install.packages("knitr")
```


# Data upload

Data extracted from noise exposure reports were recorded in Microsoft Excel spredsheet (Excel). Basic data preparation (i.e., cleaning, clustering, verification of missing data) was done in Excel. Final steps before analysis the data in R was to save the spreadsheet in comma separated format (.csv). The file was saved by the name "noiseprof.csv". Due to confidentiality, the raw data file cannot be attached in this document, however the codes used can be generalized for other applications.

Step 1: Upload dataset ("noiseprof.csv") into R
To upload data frame in .csv use the following function and assign to a new variables (optional). There are many other ways to load data from .csv as explained in YouTube
```{r}
noisedf <- read.csv(noiseprof.csv, stringsAsFactors = TRUE, na.strings = c("", "NA"))
```

Another way you may consider is to upload your dataset in Excel native file (.xlsx or .xls). In doing so, you are required to install "XLConnect" package from CRAN database. This is a matter of preference and just provide you with the alternative way to upload your dataset into R for further analysis.
```{r}
install.packages("XLConnect") #to install package within R environment
library("XLconnect") #required in every new session to enable functions within the package
```

Step 2: Explore "noiseprof.csv" 
The following are functions which can be used to explore uploaded data frame. Use help function within R to understand the arguments / values required to use respective functions
```{r}
dim() # to have overview of the dataframe (row, column)
head() # use together with tail() to check consistentcy of the dataframe
tail() # use together with head() to check consistentcy of the dataframe
str() # use to understand structure / category of the variables. here we can identify appropriate method for treatment / analyses (i.e., variables numericals recognizes as factor and vise versa)

```

# List of variables under study
```{r}
names(noisedf)
```

Selection of specific variables of interest will be explained under data preparation for statistical analysis section



# Data exploration and treatment for statistical analysis

A) Exploration 

0. Read table
```{r}
noisedf1 <- tbl_df(noisedf) # pass your data frame using tbl_df function (e.g., noisedf). Check as.tibble()
```


1. Summary
```{r}
summarize()

summarize(by_bu, mean(oeltype)) # as example

summary(split(noisedf1, noisedf1$bu)) # provide summary by business

```


2. Investigation by subset (e.g., business, seg)
```{r}
which() # to locate missing data, return TRUE values to logical test pass to the function

noisedf1 [which(noisedf1$oel >= 140), ] # as example to look for observations within noisedf1 where oeltype = peak_dB, oel => 140, which()only find TRUE value

noisedf1[noisedf1$oel >= 140, ] # same as above

noisedf1 [noisedf1$variables] # can use to access variables of interest, and apply function as desired.

noisedf1 [noisedf1$oel >= 140, ] # as example: access all meet the criteria specified in the filter

```


3. Graphic

The following functions are useful in exploring the data set:
```{r}
library(ggplot2)

boxplot(noisedf1[noisedf1$oel & noisedf1$seg == "shiftProd"])

qplot(seg, peak, data = noisedf1)

```

B) Treatment / Wrangling

```{r}

library(dplyr)

gsub() # use to convert factor to integer by replacing characters (e.g., "$", "%"")

select() # to select variable (column) of interest within the dataframe

noisepeak <- select(noisedf1, bu, oeltype) #as example

noisemax <- select(noisedf1, bu, oeltype) #as example

noisetwa <- select(noisedf1, bu, oeltype) #as example

filter() # to select subset of column

peak_bu <- filter(noisepeak, oeltype == "peak_db") #as example

max_bu <- filter(noisepeak, oeltype == "stel_dba") #as example

group_by() #break up your dataset into groups of rows based on the values of one or more variables.

group_by(noisedf, oeltype = peak_dB) # as example

```


C) Missing data

0. Identify

The are many approach in R to identify missing data or NA values within the dataset. The idea is to find and explore appropriate treatement to the missing data or NA as described in the following steps. below are several codes relevant to the process.
```{r}

noisedf1[!complete.cases(noisedf1), ] # searching for NA / empty cells or incomplete rows across dataset
noisedf1 [!complete.cases(noisedf1), ]
noisedf1[is.na(noisedf1$NRReffective), ] # this will highlight the values == "NA" in NRReffective column of noisedf1 dataset
noisedf1[is.na(noisedf1$NRReffective) & noisedf1$bu == "dsgas", ] # similar to previous, this add more criteria to the filter ([]) - "in dsgas"



```

1. Verification - for hearing protection type / NRR

2. Median imputation - for hearing protection type / NRR
```{r}
med_nrr_wu <- median(noisedf1[noisedf1$seg == "shiftprodtech",  ], na.rm = TRUE) #?median

noisedf1[is.na(noisedf1$NRReffective), ] <- med_nrr_wu #the idea is to find the NA and replace with specific values

```


3. Removal - when critical variables were absent (e.g., exposure level, % dose)
```{r}
# always back up data set prior to removal of any observations

noisedf_backup <- noisedf1

noisedf2 <- noisedf1 [!is.na(noisedf1), ] # opposite of is.na - identifying row (observation with NA values)


```



# Descriptive Data Analysis

1. Descriptive summary

```{r}
hist()
qplot(peak, data = noised1, color = bu)

xyplot (peak ~ bu | seg, data = noisedf1) # explore lattice system

summary(noisedf1$dose) # summary for specific variables

quantile()

percent_rank()

boxplot()
qplot(bu, peak, data = noisedf1, geom = "boxplot") # boxplot of peak by business
qplot(seg, peak, data = noisedf1, geom = "boxplot") # boxplot of peak by business
```


# Statistical Inferences

```{r}
norm()
t.test()

```


# Exploratory Data Analysis (Visual)

this sections will provide codes to produce graph / diagram in the text

1. boxplot
boxplot(peak~seg, noisedf1)
d <- noisedf 1
d + geom_jitter + geom_boxplot (aes(x = bu, y = oel))

2. scatterplot

with(noisedf1, plot(stel, seg))
qplot(displ, hwy, data = mpg, facets = .~ drv) # display by oeltype?
g <- ggplot(mpg, aes(displ, hwy))
g + geom_point() + geom_smooth() #draw confidence band within the scatter plot

# there are several plotting system - for consistency in looks and feel, ggplot is main choice unless other plotting system is better in representing the data.


# Learn R using Swirl
```{r}
library(swirl)
install_course("Getting and Cleaning Data")
swirl()

```

# Version control using GitHub

```{r}
#You cannot push anything that hasn't been committed yet. The order of operations is:

#1. Make your change.
# git add - this stages your changes for committing
# git commit - this commits your staged changes locally
# git push - this pushes your committed changes to a remote

```

# Reference

1. Swirl
2. R
3. R Studio

# Other useful codes

```{r}
getwd() # to confirm working directory for the project

setwd() # to set working directory when required

setwd("/Users/alifffirdausmy/Documents/GitHub/noiseproR") # working directory for this project

noisedf_0 <- noisedf # backup of original dataset
```


